---
title: "Seismic safety performance of California hospitals in 2020"
author: "Shawn Ye"
output: 
    html_document:
        toc: TRUE
        toc_float: TRUE
---
<br>

# Introduction
In the past, many hospitals in California were damaged and evacuated due to earthquakes. Some even collapsed and led to death of patients. All told, the quake had placed a heavy burden on California’s hospital system. Therefore, the state officials demand that by 2030, the hospitals not only need to remain standing but can also continue to operate after the earthquakes. 

The primary interests of this project are to:  
- Identify the best general acute care hospitals in California with reference to Structural Performance Category (SPC) and Non-structural Performance Category (NPC).       
- Look at the distribution of hospitals in CA.            
- Look at the overall seismic safety performance of California hospitals in 2020.

* SPC: ratings range from 1 to 5 with SPC 1 assigned to buildings posing significant risk of collapse following a strong earthquake and SPC 5 assigned to buildings reasonably capable of providing services to the public following a strong earthquake
* NPC: ratings range from 1 to 5 with NPC 1 assigned to buildings where the safe and orderly evacuation following a strong earthquake cannot be assured and NPC 5 assigned to buildings capable of continued operation for 72 hours without any power, water and sewer services following a strong ground motion

<br>

# Interactive visualization
```{r, include=FALSE}
library(data.table)
library(dtplyr)
library(dplyr)
library(lubridate)
library(leaflet)
library(ggplot2)
library(knitr)
library(kableExtra)
library(lubridate)
library(qwraps2)
library(plotly)
```

```{r, include=FALSE}
# read in the datasets
dt <- data.table::fread("/Users/SF/Desktop/Study/PM566/ca-oshpd-gachospital-buildingspcnpcratings-09252020.csv")

# check the dimensions, headers and footers
dim(dt)
head(dt)
tail(dt)

# check the variable names and variable types
str(dt)

# check the key variables.
colnames(dt)[colnames(dt) == 'County Code'] <- 'county'
colnames(dt)[colnames(dt) == 'SPC Rating'] <- 'SPC'
colnames(dt)[colnames(dt) == 'OSHPD NPC Rating'] <- 'NPC'

table(dt$SPC)
(sum(dt$SPC == 'N/A') + sum(dt$SPC == '-1')) / length(dt$SPC)
dt <- dt[which(dt$SPC != 'N/A')]
dt <- dt[which(dt$SPC != '-1')]
dt$SPC <- substr(dt$SPC, 1, 1)
table(dt$SPC)

table(dt$NPC)
(sum(dt$NPC == 'N/A') + sum(dt$NPC == 'NYA'))/ length(dt$NPC)
dt <- dt[which(dt$NPC != 'N/A')]
dt <- dt[which(dt$NPC != 'NYA')]
table(dt$NPC)

dt$SPC <- as.integer(dt$SPC)

# only analyze those in-service buildings
table(dt$`Building Status`)
dt <- dt[which(dt$`Building Status` == 'In Service')]
table(dt$`Building Status`)

dt$NPC <- as.integer(dt$NPC)

# check the number of counties and facilties
length(unique(dt$county))
length(unique(dt$`Facility Name`))
dt$counid <- substr(dt$county, 1,2)
dt$counname <- substr(dt$county, 5,100)

# check the range of lattitude and longitude.
summary(dt$Latitude)
summary(dt$Longitude)
```

```{r, include=FALSE}
# the number of buildings of every facility
ctf_build <- table(dt$`Facility Name`)
ctf_build_10 <- as.data.frame(head(sort(ctf_build, decreasing = TRUE),10))
ctf_build_10


# the average SPC of every facility
meanf_spc <- sort(tapply(dt$SPC,dt$`Facility Name`,mean), decreasing = TRUE)
meanf_spc5 <- as.data.frame(meanf_spc[which(meanf_spc==5)])
topspc <- rownames(meanf_spc5)
  
# the average NPC of every facility
meanf_npc <- sort(tapply(dt$NPC,dt$`Facility Name`,mean), decreasing = TRUE)
meanf_npc5 <- meanf_npc[which(meanf_npc==5)]
topnpc <- rownames(meanf_npc5)

#find the best facility
intersect(topspc, topnpc)

# the average SPC of every county
meanc_spc <- as.data.frame(sort(tapply(dt$SPC,dt$county,mean), decreasing = TRUE))
meanc_spc$county_name <- rownames(meanc_spc)
colnames(meanc_spc) <- c('Mean SPC', 'county_name')

# the average NPC of every county
meanc_npc <- as.data.frame(sort(tapply(dt$NPC,dt$county,mean), decreasing = TRUE))
meanc_npc$county_name <- rownames(meanc_npc)
colnames(meanc_npc) <- c('Mean NPC', 'county_name')

# find the counties with best performance
newtb <- merge(meanc_spc, meanc_npc, by= 'county_name')
newtb$perf <- (newtb$`Mean SPC` + newtb$`Mean NPC`)/2
county_top <- head(newtb[order(newtb[,4], decreasing = TRUE),],55)
county_top <- as.data.table(county_top)

# the number of facilities of LA
countyf <- dt[,c("county","Facility Name")]
countyf <- countyf[!duplicated(countyf),]
tb <- as.data.frame(table(countyf$county))
```

## Comparison of buildings and facilities
### Barchart1 1: Top 10 facilities with most buildings
```{r, echo=FALSE}
p <- ggplot(ctf_build_10, mapping = aes(x = substr(as.character(Var1),1,50), y = Freq)) + geom_bar(stat = 'identity',fill = 'salmon2') + xlab('hospital name') + theme(axis.text.x = element_text(angle = 45, hjust = 0.5, vjust = 0.5))
ggplotly(p)
```
Sharp Memorial Hospital, Torrance Memorial Medical Centre, Rady Children’s Hospital – San Diego and St. Joseph Hospital are the top four hospitals with most buildings in California. 

<br>

### Barchart 2: Top 15 counties with most facilities
```{r, echo=FALSE}
ctc_faci_10 <- head(tb[order(tb[,2],decreasing = TRUE),],15)
p <- ggplot(ctc_faci_10, mapping = aes(x = substr(as.character(Var1),1,30), y = Freq)) + geom_bar(stat = 'identity',fill = 'skyblue3') +xlab('county_ID') + theme(axis.text.x = element_text(angle = 45, hjust = 0.5, vjust = 0.5))
ggplotly(p)
```

<br>

## Barchart 3: 15 counties with the fewest facilities
```{r, echo=FALSE}
ctc_faci_10d <- head(tb[order(tb[,2],decreasing = FALSE),],15)
pd <- ggplot(ctc_faci_10d, mapping = aes(x = substr(as.character(Var1),1,30), y = Freq)) + geom_bar(stat = 'identity',fill = 'skyblue3') +xlab('county_ID') + theme(axis.text.x = element_text(angle = 45, hjust = 0.5, vjust = 0.5))
ggplotly(pd)
```
Los Angles has the most general acute care facilities. However, there are many counties that lack hospitals. 

<br>

## Scatterplot 1: SPC and NPC scores by county
```{r,echo=FALSE,warning=FALSE}
county_top %>% 
           plot_ly(x = ~`Mean SPC`, y = ~`Mean NPC`, 
                   type = 'scatter',
                   mode = 'markers',
                   color = ~county_name
                   )
```
In general, a majority of counties have an average NPC between 2 and 3 as well as an average SPC between 3 and 4.

<br>

## Leaflet 1: Hospital distribution in CA
```{r,echo=FALSE}
library(RColorBrewer)
colourCount = length(unique(dt$counname))
getPalette = colorRampPalette(brewer.pal(9, "Set1"))
pal <- colorFactor(
  palette = getPalette(colourCount),
  domain = dt$counname
)
leaflet(dt) %>%
  addProviderTiles('OpenStreetMap') %>%
  addCircles(lat=~Latitude, lng=~Longitude, radius = 0.01,color = ~pal(counname)
  )
```
A majority of the facilities are located around Los Angeles, Orange and San Francisco. 

<br>

## Table 1: SPC and NPC for all counties
```{r, echo=FALSE,warning=FALSE}
# summary table for scp, ncp for all counties
# kable(county_top,'html',caption = 'Table1: Ordering Seismic Safety Performance For All Counties In CA')%>%
  #kable_styling(bootstrap_options = c("striped", "hover"))%>%
  #footnote(general = "perf=(Mean SPC+Mean NPC)/2 ")
county_top$`Mean SPC` <- round(county_top$`Mean SPC`, 2)
county_top$`Mean NPC` <- round(county_top$`Mean NPC`, 2)
county_top$perf <- round(county_top$perf, 2)
library(DT)
tb <- county_top %>% select(county_name, `Mean SPC`, `Mean NPC`, perf)
datatable(tb)

```

<br>

# Full report
The link to the report is.
